---
title: "STAT4010 2021S Karte 6"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
    number_sections: no
    toc: no
    toc_depth: 2
  pdf_document:
    number_sections: no
    toc: no
    toc_depth: 2
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10, out.width = "100%")
```

# Announcement
1. The midterm scores will be uploaded once Keith has finished the adjustments. Please ask Zhang Zheng if you have any questions. I have reviewed some of his gradings, which are fine in general. I will elaborate more in the tutorial (and in the review again before final).

2. The mathematical stuffs in Ch6 will not be tested. However, you need to understand the concepts like difference in precision assessments between frequentist and Bayesian.

3. The assignments will be more and more computational starting from A4. Please come to the tutorial if you need help with R.

# Assignment 4
If I omit a question, that usually means the hints are sufficient so you can just refer to the Remark in the assignment.

## Ex4.1.1
Modify the implementation in [Karte 5](https://hemanlmf.github.io/teaching/stat4010/2021S/Karte-5.html):
```{r ex4.1.1, eval=FALSE}
et = function(x, posterior, support=c(0,1), n=301, alpha=0.005, plot=TRUE){
	# step 1: values of posterior at different values of theta in the support
	theta = seq(from=support[1], to=support[2], length=n)
	d = posterior(theta, x)
	d = d/sum(d)
	# step 2: find the theta that satisfy the credibility requirement 
	csd = cumsum(d)
	L = sum(csd < alpha/2)
	U = sum(csd < 1-alpha/2)
	# step 3: plot
	if (plot) {
	  plot(theta, d, type="l", lwd=2, col="red4",
	       xlab=bquote(theta),
	       ylab=bquote(pi(theta~"|"~italic(x[1:n]))))
	  abline(v=theta[L:U], col="pink")
	  abline(v=c(support,L,U), lty=2, lwd=.75)
	}
}
```

## Ex4.1.2
There is a new computational problem in this question. To see it, note that the posterior can be written as:
$$
f(\theta \mid x_{1:n}) \propto \theta^{\sum_{i=1}^n x_i} e^{-n\theta} \texttt{dnorm}(\theta, \texttt{mean}=0.5, \texttt{sd}=0.1) \mathbb{I}(0\leq \theta\leq 5).
$$
If we implement it directly, an obvious problem is that $\theta^{\sum_{i=1}^n x_i}$ can be extremely small or large depending on $\theta$ (e.g., $0.9^{1101}$ and $1.1^{1101}$), which is invalid in computation. The first trick that we can think of is the $\exp\{\ln(\cdot)\}$ trick; see [Karte 2](https://hemanlmf.github.io/teaching/stat4010/2021S/Karte-2.html). Then the posterior can be written as:
$$
f(\theta \mid x_{1:n}) \propto \exp\left\{ \left(\sum_{i=1}^n x_i \right) \ln \theta -n\theta \right\} \texttt{dnorm}(\theta, \texttt{mean}=0.5, \texttt{sd}=0.1) \mathbb{I}(0\leq \theta\leq 5).
$$
However, this does not solve the problem as $\left(\sum_{i=1}^n x_i \right) \ln \theta -n\theta$ is quite small:
```{r ex4.1.2_expln}
theta = seq(from=0, to=5, length=11)
1101*log(theta)-4010*theta
exp(1101*log(theta)-4010*theta)
```
To solve this problem completely, note that
$$
f(\theta \mid x_{1:n}) \propto \exp\left\{ \left(\sum_{i=1}^n x_i \right) \ln \theta -n\theta +k \right\} \texttt{dnorm}(\theta, \texttt{mean}=0.5, \texttt{sd}=0.1) \mathbb{I}(0\leq \theta\leq 5)
$$
is valid as long as $k$ is a constant that does not depends on $\theta$. Then, we can choose $k$ such that $\max_{\theta \in [0,5]} \left(\sum_{i=1}^n x_i \right) \ln \theta -n\theta +k = 0$ to normalize the density. Below is a draft of the function for the posterior:
```{r ex4.1.2_post, eval=FALSE}
post = function(theta, x){
  n = 4010 #ignore the argument x as the summary statistics are given
  s = 1101
  k = #find a reasonable k as described in the above
  exp(s*log(theta)-n*theta+k) *dnorm(theta, 0.5, 0.1) *(theta>=0) *(theta<=5)
}
```

## Ex4.2.2
Use the $\exp\{\ln(\cdot)\}$ trick.

## Ex4.3.2
The most difficult part of this question is to extract and compute $x$ according to $\texttt{racecourse}$, $\texttt{runway}$ and $\texttt{distance}$. Here I will illustrate how to extract $x$ when $\texttt{distance} = 1000$:
```{r ex4.3.2}
# Download data
id = "1t0YIve2ACvspGrQ-1htTHbhBYkjAgOh2"
data = read.table(sprintf("https://docs.google.com/uc?id=%s&export=download",
                          id),header=TRUE)
# Set conditions to filter data
distance = 1000
innerDraw = 6
cond1 = data[,"distance"]==distance
cond2 = data[,"racecourse"]=="HV"
cond3 = data[,"runway"]=="TF"
cond = cond1 &cond2 &cond3
# Filter data
dataCond = data[cond, c("race","draw","position")]
race = unique(dataCond[,"race"])
# Compute x
x = rep(NA, length(race))
for (i.race in 1:length(race)) {
  dataRace = dataCond[(dataCond[,"race"])==(race[i.race]),]
  x[i.race] = mean(dataRace[dataRace[,"draw"]<=innerDraw, "position"]) <
    mean(dataRace[dataRace[,"draw"]>innerDraw, "position"])
}
c(length(x), sum(x), sum(x)/length(x))
```