---
title: "STAT3005 2020F A Question on Monte Carlo Experiment"
author: "LEUNG Man Fung, Heman"
output:
  pdf_document:
    toc: no
    toc_depth: 2
  html_document:
    df_print: paged
    toc: no
    toc_depth: 2
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question
How to implement a Monte Carlo experiment in R?

# Answer
The flow of a Monte Carlo experiment is as follows:

1. Generate data from some probabilistic models
2. Perform a deterministic computation on the data
3. Aggregate the results

Applying this flow on approximating the value of $\pi$ (a graphical illustration can be found on [Wikipedia](https://en.wikipedia.org/wiki/Monte_Carlo_method#/media/File:Pi_30K.gif)), the flow becomes:

1. Generate $U_1, U_2 \sim U(0,1)$
2. If $\sqrt{U_1^2+U_2^2} \le 1$, then set $V_i = 1$. Otherwise set $V_i = 0$. The idea here is to check whether the point falls inside 1/4 of a unit circle
3. Repeat step 1 and 2 for $n$ times. The approximated value of $\pi$ is $4n^{-1} \sum_{i=1}^n V_i$

We try to implement the above in R:
```{r, eval=T}
n = 10000
V = vector(length=n)
# 3. Repeat step 1 and 2 for n times
for (i in 1:n)
{
  # 1. Generate U1, U2
  U = runif(2)
  # 2. Check whether the point falls inside 1/4 of a unit circle
  if (sqrt(U[1]^2+U[2]^2) <= 1)
  {
    V[i] = 1
  } else
  {
    V[i] = 0
  }
}
4*sum(V)/n # Approximated value of pi
```

We can further try different radius to check the validity, which is similar to the idea of *n.all* or *delta.all* in this course. The flow becomes:

1. Generate $U_1, U_2 \sim U(0,r_j)$
2. If $\sqrt{U_1^2+U_2^2} \le r_j$, then set $V_i^{(j)} = 1$. Otherwise set $V_i^{(j)} = 0$. The idea here is to check whether the point falls inside 1/4 of a circle with radius $r_j$
3. Repeat step 1 and 2 for $i=1,\dots,n$ and $j = 1,\dots,m$. The approximated values of $\pi$ are $\hat{\pi}^{(j)} = 4n^{-1} \sum_{i=1}^n V_i^{(j)}$

We try to implement the above in R. Note that some parts are extended compared with the last snippet.

```{r, eval=T}
n.rep = 10000
rad.all = c(1, 1.5, 2, 2.5, 3)
n.rad = length(rad.all)
V = matrix(nrow=n.rep, ncol=n.rad)
rownames(V) = paste0("i.rep=",1:n.rep)
colnames(V) = paste0("rad=",rad.all)
# 3. Repeat for different radius
for (i.rad in 1:n.rad)
{
  # 3. Repeat for different replications
  for (i.rep in 1:n.rep)
  {
    # 1. Generate U1, U2 according to radius r
    r = rad.all[i.rad]
    U = runif(2, 0, r)
    # 2. Check whether the point falls inside 1/4 of a circle with radius r
    V[i.rep, i.rad] = (sqrt(sum(U^2)) <= r)
  }
}
4*colSums(V)/n # Approximated value of pi
```

The output conveys two messages. Firstly, the method seems to be working for different radius. Second and more importantly, the approximated value when radius is 1 probably differs from previous one. This is because step 1 of Monte Carlo experiment involves randomness and so error is likely to exist. To make the results reproducible, we can leverage on the *set.seed* function. For instance,

```{r, eval=T}
for (i.rad in 1:n.rad)
{
  set.seed(i.rad) # control randomness for reproducible results
  for (i.rep in 1:n.rep)
  {
    r = rad.all[i.rad]
    U = runif(2, 0, r)
    V[i.rep, i.rad] = (sqrt(U[1]^2+U[2]^2) <= r)
  }
}
4*colSums(V)/n
```

The result can be reproduced when we run the same snippet:
```{r, eval=T}
for (i.rad in 1:n.rad)
{
  set.seed(i.rad) # control randomness for reproducible results
  for (i.rep in 1:n.rep)
  {
    r = rad.all[i.rad]
    U = runif(2, 0, r)
    V[i.rep, i.rad] = (sqrt(U[1]^2+U[2]^2) <= r)
  }
}
4*colSums(V)/n
```

We can also report to relative error so that it is more readable:
```{r, eval=T}
4*colSums(V)/n/pi-1
```

Finally, we comment on several R functions that we have used for simulation so far:

1. *array*: it is just a multi-dimensional "storage space" for our simulated values. When we try different radius above, we change *V* from 1D vector to 2D matrix during initialization
2. *apply*: it is a functional programming device for looping. If we are not familiar with it, we can always implement the same thing using traditional device like for loop