---
title: "STAT3005 2020F A Short Note on Tips for R"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 2
  pdf_document:
    number_sections: yes
    toc: no
    toc_depth: 2
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(microbenchmark)
```

# Introduction
This short note aims to give some practical tips for R as compared with some common languages that I know. Please follow a more structured guide if you want a reference on R. I have the following recommendations:

- [Notes of RMSC1101](https://sites.google.com/site/kwchankeith/teaching/r1101/r1101-2021Fall)  
These notes are relatively concise. Read it if you are completely new to R  
- [R Programming for Data Science](https://bookdown.org/rdpeng/rprogdatascience/)  
This book is fairly complete for beginners. Use it as a reference as not everything inside are relevant to our course  
- [Advanced R](https://adv-r.hadley.nz/)  
This book is, as its name suggests, an advance treatment. Read it if you would like to know more as an experience user  

# Basic
## The Survival *?*
We can inspect the documentation of everything in R. For example,
```{r, eval=F}
?mean
?var
```
We can even inspect more elementary commands (though we will need to add quotation marks for those):
```{r, eval=F}
?"+"
?"for"
?"?"
```
Looking at the documentation is helpful when we are not sure about how to use some functions (e.g., from a new package). The help manual typically tells us how to use the function, its expected output and provides some examples that we can learn from.

## Long Statements
(If you just want a quick fix, you can toggle the soft-warp. For R-studio, go to Tools > Global Options > Code and check "Soft-warp R source files".)

Sometimes we may have some long statements that exceed the width of editor. This can be inconvenient when we read the same code later. As R do not have end statement like ";" in other languages, whenever we break down a long statement we have to make sure that it is not ended yet. Here is an example:
```{r}
# Correct multiline statement
x = 1 +2 +3 +4 +5 +
  6 +7 +8 +9
x # 45
# Incorrect multiline statement
y = 1 +2 +3 +4 +5
  +6 +7 +8 +9
y # not 45
```
There is no error for the second case because "+6 +7 +8 +9" is also a valid statement. However, the code is not working as we want.

## Subsetting
In general, we subset vector, matrix or vector-like data structure with "[". For matrix, the order of indexes should be row and then column (same as the tradition in Math). Note that the index in R starts from 1 unlike some other languages. Here are some examples:
```{r}
# Vector
x = 1:9
x[5] # extract single element
x[5:7] # extract multiple elements
x[-1] # drop single element
x[-c(3,5)] # drop multiple elements
# Matrix
y = matrix(1:9, nrow=3, ncol=3)
y
y[1:2, 2:3] # subset a matrix
```
It is also possible to use string as the index provided that the data structure has names:
```{r}
# Name the previous matrix
colnames(y) = c("S","T","A")
y[1:2, c("S","A")] # subset a matrix with string
y[1:2, c(1,3)] # same as above
```
For list-like data structure, we extract an element with "[[" (or "$" provided that the element is named). For example:
```{r}
# List
z = list(code=3005, subject="STAT", school="CUHK")
z$code
z[["subject"]]
z["school"] # this returns a list
```
List can be used to store heterogeneous data so it is less often used as a data structure in our course.

## Looping
There are several common ways to do iterations in R:

1) Traditional control flow such as *for*, *while* and *repeat*
2) Functional programming such as *apply* and *lapply*
3) Vectorization

For beginners, you may stick to traditional control flow until you are more familiar with R. We will discuss vectorization if we have time.

## Order of Arguments
When we call a function, we do not need to follow the order of arguments if we name the arguments. For example,
```{r}
# cov(x, y = NULL, use = "everything",
#    method = c("pearson", "kendall", "spearman"))
a = c(1:10)
b = c(11:20)
cov(method="kendall", use="everything", x=a, y=b)
cov(a, b, "everything", "kendall") # same as above
```
In practice, we can use this fact to skip specifying some arguments which have default value that we are fine with. For example,
```{r}
cov(a, b, method="kendall")
```

# Intermediate
## Ternary If Operator
In some common programming languages, *?:* is provided as a ternary if operator. For instance, the statement
```{c, eval=F}
a ? b : c
```
means "if a then b otherwise c". However, *?* is the help operator, which is shown in [The Survival *?*], so this is not true in R. Commonly, people think that *ifelse* is the ternary if operator but this is not accurate. The correct ternary if is *\`if\`* while *ifelse* is its vectorized version. We try to compare their speed with the microbenchmark library:
```{r}
U = runif(1000)
f_if = function(U, n=length(U)) # use `if`
{
  out = vector(length=n)
  for (i in 1:n) # can skip {} if only 1 line
    out[i] = `if`(U[i]<0.5, 0, 1)
  out
}
f_eif = function(U, n=length(U)) # use ifelse
{
  out = vector(length=n)
  for (i in 1:n) # can skip {} if only 1 line
    out[i] = ifelse(U[i]<0.5, 0, 1)
  out
}
f_eif_vec = function(U, n=length(U)) # use ifelse in a vectorized way
{
  out = ifelse(U<0.5, 0, 1)
  out
}
all.equal(f_if(U), f_eif(U), f_eif_vec(U)) # sanity check
# evaluate execution time systematically
print(microbenchmark(f_if(U), f_eif(U), f_eif_vec(U), times=50))
```
There are two messages here. First, *\`if\`* is the correct ternary if operator since it is much faster when a single element is concerned. Second, vectorization can speed up the implementation in R. The general idea is to use vectors instead of single elements as inputs into a vectorized function.

## Pre-allocation
In simulations, we frequently pre-allocate the storage *out* in advance of the main loops. The reason is that dynamic memory allocation is usually costly in R. To see this, we try a simple example:
```{r}
n = 1000
f_pre = function(n) # pre-allocate out
{
  out = vector(length=n)
  for (i in 1:n)
    out[i] = i
  out
}
f_dyn = function(n) # combine elements dynamically
{
  out = 1
  for (i in 2:n)
    out = c(out, i)
  out
}
all.equal(f_pre(n), f_dyn(n))
print(microbenchmark(f_pre(n), f_dyn(n), times=50))
```
By adjusting the value of *n*, we can see that dynamic memory allocation is not of constant time in this way. Therefore, we should try our best to pre-allocate the necessary storage to avoid performance issues.

## Looping with Functionals
In [Looping], I have recommended stick to traditional control flow. The rationale is that functional loops can usually be rewritten as traditional loops with similar performance. Consider the following:
```{r}
x = matrix(runif(5000), nrow=1000, ncol=5)
f_for = function(x, n=nrow(x)) # row sum with for loop
{
  out = vector(length=n)
  for (i in 1:n)
    out[i] = sum(x[i,])
  out
}
f_apply = function(x) # row sum with apply
{
  out = apply(x, 1, sum)
  out
}
f_vec = function(x) # vectorized row sum
{
  out = rowSums(x)
  out
}
all.equal(f_for(x), f_apply(x), f_vec(x))
print(microbenchmark(f_for(x), f_apply(x), f_vec(x), times=50))
```
In this simple case, *for* is marginally faster than *apply* while the vectorized *rowSums* is much faster than both. When should we use functional loops in practice? There are several cases:

1) Readability: functional loops result in neater code in general
2) Design pattern: functional style of programming may be preferred

## Sourcing
The concept of package in R is similar to the concept of library in the other programming languages. However, creating packages to manage our code can be time consuming. A quicker way to save and reuse common R code is through the function *source*. For example,
```{r}
x = "stat"
source("example.R") # the code inside covered last line
x
```
We can see that the value of *x* changes because it was "covered" by the code in example.R, which consists of
```{r, eval=F}
# Content of example.R
x # this will not be printed
x = 3005
```
There are several things to note:

1) Intuitively, *source* "import" R code by running every line. If global variables/operations are used in the to-be-imported file, they might affect the result unexpectedly. Put only functions in the to-be-imported file to avoid this issue  
2) By the above example, we can see that some operations might not have effect. For example, calling *x* (the second line in example.R) did not result in auto-printing the value of *x*, which was "stat". Read the documentation for details about the behavior of *source*  
3) It is possible to import global code chunks separately if we use R markdown. See [this](https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html). However, we still need to name the chunks so that we can pinpoint which one to be called  

## Command Line
While R studio or other IDE is more user friendly, running R code in them is usually slower than running via command line. One of the reasons is the dynamic interaction with the environments. In R studio, we can inspect the variables usually by clicking their names in the environment panel. However, this convenience comes at cost of tracing and showing the memories associated with the names. Therefore, if execution speed is of concern after the development stage, we should run our R code via terminal directly (e.g., using *source*)

# Advance
## Vectorization
Vectorization, or array programming, refers to applying operations to the whole array instead of its individual elements one by one. It is an important concept in R because suitable vectorization can speed up R program a lot in practice. We have presented two simple examples in [Ternary If Operator] and [Looping with Functionals]. Here we try to give clearer picture via another example.

In some common programming languages, a sum of vector elements function is usually implemented as follows:
```{r}
sum_loop = function(v)
{
  s = 0
  for (i in 1:length(v))
    s = s +v[i]
  s
}
```
We compare this with the built-in *sum* function, which is vectorized:
```{r}
x = 1:1000
all.equal(sum_loop(x), sum(x))
print(microbenchmark(sum_loop(x), sum(x), times=50))
```
Obviously, the built-in *sum* is much faster. Nevertheless, if we think about this and the previous two examples carefully, we will realize that we cannot write vectorized R functions ourselves in some sense. The *ifelse*, *rowSums* and *sum* functions are vectorized because they are built-in with highly optimized compiled code (probably in C). It would be difficult to write a vectorized R function without using them because we would not deal with those (probably C) code in general.

In other words, the building blocks of vectorization in R are usually the functions provided in system library (take a look at the package panel in R studio). As long as we operate with arrays directly with suitable built-in functions, we are doing vectorization in R in a narrow sense. To illustrate different possibilities, we try to give a more complex example. Consider a data series with missing data:
```{r}
set.seed(3005) # to ensure A will not change
n = 10000
X = round(runif(n), 3)
A = sort(sample(1:n, n/2, F))
head(X, 10) # full data series
head(X[A], 10) # observed data series
head(A, 10) # index of observed data
```
We are interested in constructing an index series of last observed data, i.e.,
$$
1, 1, 1, 4, 5, 6, 7, 8, 8, 8,\ldots
$$
A naive implementation is
```{r}
f_naive = function(A, n)
{
  out = rep(0, n)
  j = 1
  for (i in A[1]:n)
  {
    if (j == n/2) next
    if (i == A[j+1]) j = j +1
    out[i] = A[j]
  }
  out
}
```
A marginally better implementation is
```{r}
f_naive2 = function(A, n)
{
  out = rep(0, n)
	out[A] = A
  for (i in A[1]:n)
    out[i] = `if`(out[i]==0, out[i-1], out[i])
  out
}
```
A vectorized implementation that uses *cumsum* is
```{r}
f_vec = function(A, n)
{
  I = rep(0, n)
  I[A] = 1 # indicator of observed data
  I = cumsum(I) # cumsum is vectorized
  out = A[I]
  out
}
```
Finally, we compare their speed below:
```{r}
all.equal(f_naive(A,n), f_naive2(A,n), f_vec(A,n))
print(microbenchmark(f_naive(A,n), f_naive2(A,n), f_vec(A,n), times=50))
```

## Environments

## Rcpp