---
title: "STAT3005 2020F A Short Note on Tips for R"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: 2
  pdf_document:
    number_sections: yes
    toc: no
    toc_depth: 2
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
Please follow a more structured guide like Keith's notes if you have no programming experience at all. This short note aims to give some practical tips in R as compared with some common languages that I know.

# Basic
## The Survival "?"
We can inspect the documentation of everything in R. For example,
```{r, eval=F}
?mean
?var
```
We can even inspect more elementary commands (though you will need to add quotation marks for those):
```{r, eval=F}
?"+"
?"for"
```
Looking at the documentation is helpful when we are not sure or forget about how to use a function. The help manual typically tells us how to use the function, its expected output and provides some examples that we can learn from.

## Long Statement
(If you just want a quick fix, you can toggle the soft-warp. For R-studio, go to Tools > Global Options > Code and check "Soft-warp R source files".)

Sometimes we may have some long statements that exceed the width of editor. This can be inconvenient when we read the same code later. As R do not have end statement like ";" in other languages, whenever we break down a long statement we have to make sure that it is not ended yet. Here is an example:
```{r, eval=T}
# Correct multiline statement
x = 1 +2 +3 +4 +5 +
  6 +7 +8 +9
x # 45
# Incorrect multiline statement
y = 1 +2 +3 +4 +5
  +6 +7 +8 +9
y # not 45
```
There is no error for the second case because "+6 +7 +8 +9" is also a valid statement. However, the code is not working as we want.

## Subsetting
In general, we subset vector, matrix or vector-like data structure with "[". For matrix, the order of indexes should be row and then column (same as the tradition in Math). Note that the index in R starts from 1 unlike some other languages. Here are some examples:
```{r, eval=T}
# Vector
x = 1:9
x[5] # extract single element
x[5:7] # extract multiple elements
x[-1] # drop single element
x[-c(3,5)] # drop multiple elements
# Matrix
y = matrix(1:9, nrow=3, ncol=3)
y
y[1:2, 2:3] # subset a matrix
```
It is also possible to use string as the index provided that the data structure has names:
```{r, eval=T}
# Name the previous matrix
colnames(y) = c("S","T","A")
y[1:2, c("S","A")] # subset a matrix with string
y[1:2, c(1,3)] # same as above
```
For list-like data structure, we extract an element with "[[" (or "$" provided that the element is named). For example:
```{r, eval=T}
# List
z = list(code=3005, subject="STAT", school="CUHK")
z$code
z[["subject"]]
z["school"] # this returns a list
```
List can be used to store heterogeneous data so it is less often used in this course.

# Looping
There are several common ways to do iterations in R:

1) Traditional control flow such as "for", "while" and "repeat"  
2) Functional programming such as "apply" and "lapply"  
3) Vectorization  

For beginners, you may stick to traditional control flow until you are more familiar with R. We will discuss vectorization if we have time.

# Order of Arguments
When we call a function, we do not need to follow the order of arguments if we name the arguments. For example,
```{r, eval=T}
# cov(x, y = NULL, use = "everything",
#    method = c("pearson", "kendall", "spearman"))
a = c(1:10)
b = c(11:20)
cov(method="kendall", use="everything", x=a, y=b)
cov(a, b, "everything", "kendall") # same as above
```
In practice, we can use this fact to skip specifying some arguments which have default value that we are fine with. For example,
```{r, eval=T}
cov(a, b, method="kendall")
```