---
title: "RMSC5102 2020S Exercise for General Simulation"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
    toc: no
    toc_depth: '2'
  pdf_document:
    number_sections: yes
    toc: no
    toc_depth: 2
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Random variable generation
## Exercise
Consider the quadratic random variable $X$ with pdf:
$$
\begin{aligned}
f_X(x) &= 
\left\{
	\begin{array}{ll}
		\frac{3}{64} (x-2)^2 &, x \in [-2,2] \\
		0 &, x \notin [-2,2]
	\end{array}
\right. \\
\end{aligned}
$$
Propose an inverse transform algorithm and acceptance-rejection algorithm using $Y \sim N(0,1)$ to generate $X$  
(modified from RMSC4001 2014S final Q3)

## Solution
Note that for $x \in [-2,2]$,
$$
\begin{aligned}
F_X(x) &= \int_{-2}^x \frac{3}{64} (t-2)^2 dt \\
&= \left[ \frac{1}{64} (t-2)^3 \right]_{-2}^x \\
&= \frac{1}{64} (x-2)^3 +1 \\
U = \frac{1}{64} (X-2)^3 +1 \implies (X-2)^3 &= 64U-64 \\
X &= (64U-64)^{\frac{1}{3}} +2
\end{aligned}
$$
Hence the inverse transform algorithm is as follows:  
1) Generate $U \sim U(0,1)$  
2) Set $X = (64U-64)^{\frac{1}{3}} +2$

We verify the algorithm with the following R implementation:
```{r q1a}
n = 10000
U = runif(n)
X = -(64-64*U)^(1/3)+2 #R cannot handle cube root of negative
hist(X, freq=F)
px = seq(-2,2,by=0.01)
py = 3/64*(px-2)^2
lines(px, py, col="blue") #add target pdf of X
```

For the acceptance-rejection algorithm, note that
$$
\begin{aligned}
f_Y(y) &= \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2} y^2} \\
\frac{f_X(y)}{f_Y(y)} &=
\left\{
	\begin{array}{ll}
		\frac{3\sqrt{2\pi}}{64} (y-2)^2 e^{\frac{1}{2} y^2} &, y \in [-2,2] \\
		0 &, y \notin [-2,2]
	\end{array}
\right. \\
c = \max_y \frac{f_X(y)}{f_Y(y)} &= \frac{f_X(-2)}{f_Y(-2)} = \frac{3 e^2 \sqrt{2\pi}}{4}
\end{aligned}
$$
since $(y-2)^2$ is decreasing for $y \in [-2,2]$ while $e^{\frac{1}{2} y^2}$ is maximized at $y=-2$ and $y=2$. Hence the algorithm is as follows:  
1) Generate $Y \sim N(0,1)$  
2) If $Y \notin [-2,2]$, return to step 1 (due to different support)  
3) Generate $U \sim U(0,1)$  
4) If $U \leq \frac{1}{c} \cdot \frac{f_X(Y)}{f_Y(Y)} = \frac{1}{16} (Y-2)^2 e^{\frac{1}{2} Y^2-2}$, set $X=Y$. Otherwise return to step 1

We verify the algorithm with the following R implementation:
```{r q1b}
# Use result of the above
X = vector(length=n)
for (i in 1:n)
{
  repeat
  {
    repeat
    {
      Y = rnorm(1)
      if (Y >= -2 && Y <= 2)
        break
    }
    U = runif(1)
    if (U <= 1/16 *(Y-2)^2 *exp(0.5*Y^2-2))
    {
      X[i] = Y
      break
    }
  }
}
hist(X, freq=F)
lines(px, py, col="blue")
```

# Antithetic variables
## Exercise
Consider an arithmetic floating strike Asian put option which payoff is $\max(A_T-S_T, 0)$, where $S_T$ is the terminal price of the underlying asset and, for $t_{j+1}-t_j = \Delta t, j=0,1,...,n$ and $t_n=T$,
$$
A_{t_m} = \frac{1}{m} \sum_{j=1}^m S_{t_j},\quad \forall1 \leq m \leq n
$$
Suppose that $T=1, n=12, r=0.02, \sigma=0.4, S_0=100$ and the underlying asset price follows GBM. Propose an algorithm with antithetic variables to value this Asian option  
(modified from RMSC4001 2015S final Q5)

## Solution
Note that $\Delta t = \frac{1}{12}$. The algorithm is as follows:  
1) Generate $Z_j \sim N(0,1)$ for $j=1,..,n$  
2) Set $S_{t_j}^{(i)} = S_{t_{j-1}}^{(i)} \exp \left[ (r-\frac{1}{2} \sigma^2) \Delta t +\sigma \sqrt{\Delta t} Z_j \right]$ for $j=1,..,n$  
3) Set $S_{t_j}^{(m+i)} = S_{t_{j-1}}^{(m+i)} \exp \left[ (r-\frac{1}{2} \sigma^2) \Delta t -\sigma \sqrt{\Delta t} Z_j \right]$ for $j=1,..,n$  
4) Repeat step 1 to 3 for $i=1,...,m$  
5) Compute $A_T^{(i)} = \frac{1}{n} \sum_{j=1}^n S_{t_j}^{(i)}$ for $i=1,...,2m$  
6) Compute $\hat\theta_{ant} = \frac{e^{-rT}}{2m} \sum_{i=1}^{2m} \max(A_T^{(i)}-S_T^{(i)}, 0)$

We verify the algorithm with the following R implementation:
```{r q2}
rm(list=ls()) #clear variables from last question
nEstimate = 1000
nSample = 1000
estimate = matrix(nrow=nEstimate, ncol=2)
colnames(estimate) = c("standard", "antithetic")
n = 12
dt = 1/n
r = 0.02
sigma = 0.4
thetaH.std = function(m){
  S = matrix(100, nrow=m, ncol=n+1) #set S0=100
  for (i in 1:n)
  {
    Z = rnorm(m)
    S[,i+1] = S[,i]*exp((r-0.5*sigma^2)*dt+sigma*sqrt(dt)*Z)
  }
  A = rowMeans(S[,-1]) #remove S0 in rowMeans
  return(exp(-r)*mean(pmax(A-S[,n+1],0)))
}
thetaH.ant = function(m){
  S = matrix(100, nrow=m, ncol=n+1) #set S0=100
  for (i in 1:n)
  {
    Z = rnorm(m/2) #the other half are -Z for antithetic variables
    S[1:(m/2),i+1] = S[1:(m/2),i]*exp((r-0.5*sigma^2)*dt+sigma*sqrt(dt)*Z)
    S[(m/2+1):m,i+1] = S[(m/2+1):m,i]*exp((r-0.5*sigma^2)*dt-sigma*sqrt(dt)*Z)
  }
  A = rowMeans(S[,-1]) #remove S0 in rowMeans
  return(exp(-r)*mean(pmax(A-S[,n+1],0)))
}
for (i in 1:nEstimate)
{
  estimate[i,1] = thetaH.std(nSample)
  estimate[i,2] = thetaH.ant(nSample)
}
apply(estimate,2,mean)
apply(estimate,2,var)
```
We can see that the price is similar but variance is reduced

# Control variate
## Exercise
With the same setting from last question, i.e. arithmetic floating strike Asian put option, propose an algorithm using $A_{t_9}$ as control variate to value this Asian option  
(modified from RMSC4001 2015S final Q5)

## Solution
Since calculating $E(A_{t_9})$ can be difficult, we propose to estimate it via pilot simulation:  
1) Generate $Z_j \sim N(0,1)$ for $j=1,..,n$  
2) Set $S_{t_j}^{(i)} = S_{t_{j-1}}^{(i)} \exp \left[ (r-\frac{1}{2} \sigma^2) \Delta t +\sigma \sqrt{\Delta t} Z_j \right]$ for $j=1,..,n$  
3) Repeat step 1 to 2 for $i=1,...,m$ (sample size of pilot)  
4) Compute $A_{t_9}^{(i)} = \frac{1}{9} \sum_{j=1}^9 S_{t_j}^{(i)}$ for $i=1,...,m$  
5) Compute $A_T^{(i)} = \frac{1}{n} \sum_{j=1}^n S_{t_j}^{(i)}$ for $i=1,...,m$  
5) Compute pilot sample estimates $\hat\mu_{A_{t_9}}, \hat\sigma^2_{A_{t_9}}, \hat\sigma_{A_{t_9},\max(A_T-S_T,0)}$  

The main algorithm is as follows:  
1) Generate $Z_j \sim N(0,1)$ for $j=1,..,n$  
2) Set $S_{t_j}^{(i)} = S_{t_{j-1}}^{(i)} \exp \left[ (r-\frac{1}{2} \sigma^2) \Delta t +\sigma \sqrt{\Delta t} Z_j \right]$ for $j=1,..,n$  
3) Repeat step 1 to 2 for $i=1,...,m$  
4) Compute $A_{t_9}^{(i)} = \frac{1}{9} \sum_{j=1}^9 S_{t_j}^{(i)}$ for $i=1,...,m$  
5) Compute $A_T^{(i)} = \frac{1}{n} \sum_{j=1}^n S_{t_j}^{(i)}$ for $i=1,...,m$  
6) Compute $\hat{\pi}_T^{(i)} = \max(A_T^{(i)}-S_T^{(i)},0) -\frac{\hat\sigma_{A_{t_9},\max(A_T-S_T,0)}} {\hat\sigma^2_{A_{t_9}}} (A_{t_9}^{(i)} -\hat\mu_{A_{t_9}})$  
7) Compute $\hat\theta_{ctr} = \frac{e^{-rT}}{m} \sum_{i=1}^{m} \hat{\pi}_T^{(i)}$

We verify the algorithm with the following R implementation:
```{r q3}
# Use result of the above
colnames(estimate) = c("standard", "control")
# pilot simulation
mPilot = 1000
S = matrix(100, nrow=mPilot, ncol=n+1) #set S0=100
for (i in 1:n)
{
  Z = rnorm(mPilot)
  S[,i+1] = S[,i]*exp((r-0.5*sigma^2)*dt+sigma*sqrt(dt)*Z)
}
A9 = rowMeans(S[,2:10])
AT = rowMeans(S[,-1]) #remove S0 in rowMeans
muH = mean(A9)
cH = cov(A9,pmax(AT-S[,n+1],0))
vH = var(A9)
cor(A9,pmax(AT-S[,n+1],0)) #small correlation
thetaH.ctr = function(m){
  # main simulation
  S = matrix(100, nrow=m, ncol=n+1) #set S0=100
  for (i in 1:n)
  {
    Z = rnorm(m)
    S[,i+1] = S[,i]*exp((r-0.5*sigma^2)*dt+sigma*sqrt(dt)*Z)
  }
  A9 = rowMeans(S[,2:5])
  AT = rowMeans(S[,-1]) #remove S0 in rowMeans
  piH = pmax(AT-S[,n+1],0) -cH/vH *(A9-muH)
  return(exp(-r)*mean(piH)) 
}
for (i in 1:nEstimate)
{
  estimate[i,1] = thetaH.std(nSample)
  estimate[i,2] = thetaH.ctr(nSample)
}
apply(estimate,2,mean)
apply(estimate,2,var)
```
We can see that the price is similar but variance reduction effect is very small (or even none). This is actually expected from pilot simulation as correlation between $A_{t_9}$ and payoff is small. Therefore, it is important to choose a good control variate in practice  
(the original question has a better control variate but its derivation is not covered in this course)

# Importance sampling
## Exercise
Consider the simulation of a European put option on a stock with current price 100, strike price 80, interest rate 3%, volatility 40% and maturity 6 months. Assume the Black-Scholes model for the underlying stock. Provide an importance sampling algorithm based on the Laplace random variable $Y$ with pdf
$$f_Y(y) = \frac{1}{2} e^{-|y|}$$
(modified from RMSC4001 2018S final Q1)

## Solution
Let $X \sim N(0,1)$. The likelihood ratio is
$$
\begin{aligned}
\frac{f_X(y)}{f_Y(y)} &= \frac{2}{\sqrt{2\pi}} e^{-\frac{1}{2} y^2+|y|}
\end{aligned}
$$
The target function is
$$
h(X) = \max(K-S_T,0) = \max(K-S_0 e^{(r-\frac{1}{2} \sigma^2)T +\sigma \sqrt{T} X}, 0)
$$
where $K=80, S_0=100, r=0.03, \sigma=0.4, T=\frac{1}{2}$. To generate $Y \sim Laplace(0,1)$, note that
$$
\begin{aligned}
F_Y(y) &= \frac{1}{2} \int_{-\infty}^y e^{-|t|} dt \\
&= \left\{
	\begin{array}{ll}
		1-\frac{1}{2} e^{-y} &, y \geq 0 \\
		\frac{1}{2} e^{y} &, y < 0
	\end{array}
\right. \\
U = F_Y(Y) \implies Y &=
\left\{
	\begin{array}{ll}
		-\ln(2-2U) &, U \geq 0.5 \\
		\ln(2U) &, U < 0.5
	\end{array}
\right. \\
\end{aligned}
$$
Hence the algorithm is as follows:  
1) Generate $U_i \sim U(0,1)$  
2) If $U_i \geq 0.5$, set $Y_i=-\ln(2-2U_i)$. Otheriwse, set $Y_i=\ln(2U_i)$  
3) Compute $L_i = \frac{2}{\sqrt{2\pi}} e^{-\frac{1}{2} Y_i^2+|Y_i|}$  
4) Compute $h_i = \max(K-S_0 e^{(r-\frac{1}{2} \sigma^2)T +\sigma \sqrt{T} Y_i}, 0)$  
5) Repeat step 1 to 4 for $i=1,...,n$  
6) Compute $\hat\theta_{imp} = \frac{e^{-rT}}{n} \sum_{i=1}^{n} h_i \cdot L_i$

We verify the algorithm with the following R implementation:
```{r q4}
rm(list=ls()) #clear variables from last question
nEstimate = 1000
nSample = 1000
estimate = matrix(nrow=nEstimate, ncol=2)
colnames(estimate) = c("standard", "importance")
K = 80
S0 = 100
T = 1/2
r = 0.03
sigma = 0.4
thetaH.std = function(n){
  X = rnorm(n)
  ST = S0*exp((r-0.5*sigma^2)*T+sigma*sqrt(T)*X)
  return(exp(-r*T)*mean(pmax(K-ST,0)))
}
thetaH.imp = function(n){
  U = runif(n)
  Y = ifelse(U<0.5, log(2*U), -log(2-2*U))
  L = 2/sqrt(2*pi) *exp(-0.5*Y^2+abs(Y))
  h = pmax(K-S0*exp((r-0.5*sigma^2)*T+sigma*sqrt(T)*Y),0)
  return(exp(-r*T)*mean(h*L))
}
for (i in 1:nEstimate)
{
  estimate[i,1] = thetaH.std(nSample)
  estimate[i,2] = thetaH.imp(nSample)
}
apply(estimate,2,mean)
apply(estimate,2,var)
```
We can see that the price is similar but variance is reduced as Laplace r.v. has more extreme value to "escape" from the out-of-the-money situation