---
title: "RMSC5102 2020S A Question on Discrete Uniform"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question
Can we use continuous version of inverse transform for discrete uniform random variable?

# Answer
It is not appropriate in terms of algorithm. Below is an explanation:  

To use continuous version of inverse transform, note that the cdf of discrete uniform is
$$
\begin{aligned}
F(x) &= \frac{x}{m} &, x=1,...,m \\
u &= F(x) = \frac{x}{m} \\
x &= um \\
\Rightarrow F^{-1}(u) &= um
\end{aligned}
$$
There are at least two problems here:  
1) As $F(x)$ is only defined at $x=1,...,m$, the inverse function $F^{-1}(u)$ is actually not well defined. As you can try in simulation, $um$ may not be an integer  
2) Even if you round the "inverse" properly, they may fail on boundary cases. For instance, if $0 \le u \le 1$, $\lceil um \rceil$ (i.e. round up to nearest integer) will fail at $u=0$ since $\lceil 0 \rceil = 0$ may not be in the support of $X$  

Hence you should use discrete version of inverse transform for generating for discrete uniform random variable

However, if you round the the "inverse" properly, the implementation is fine in R practically because the documentation states that "runif(1,0,1)" will not return 0 or 1. This avoids the boundary cases but I do not recommend it as it depends on specification of uniform random number in the language
```{r answer}
m = 100
n = 10000
U = runif(n) #use same set of U to compare plot
X = Y = vector(length = n)
#Standard way
for (i in 1:n)
{
  for (j in 1:m)
  {
    if (U[i] > (j-1)/m && U[i] <= j/m)
    {
      X[i] = j
      break
    }
  }
}
hist(X)
#Non-standard way
for (i in 1:n)
{
  Y[i] = ceiling(U[i]*m)
}
hist(Y) #same shape as X's plot
```