---
title: "RMSC5102 2020S A Question on Monte Carlo Integration"
author: "LEUNG Man Fung, Heman"
output:
  html_document:
    df_print: paged
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question
How to use standard Monte Carlo to estimate $\theta = \int_a^b g(x) dx$ when $(a,b) \neq (0,1)$?

# Answer
When $(a,b) = (0,1)$, we generate $U \sim \textrm{Unif}(0,1)$ as the underlying random variable in standard Monte Carlo since the support of $U$ is same as interval of integration. The idea is essentially
$$
\int_0^1 g(x) dx = \int_0^1 g(u) f(u) du = \mathbb{E}\{g(U)\}, 
$$
since the pdf of $U$ is $f(u)=1$.  

Similarly, for $\theta = \int_a^b g(x) dx$, we generate $V \sim U(a,b)$. Then
$$
\int_a^b g(x) dx = (b-a)\int_a^b g(v) \frac{1}{b-a} dv = (b-a)\mathbb{E}\{g(V)\},
$$
since the pdf of $V$ is $f(v)=\frac{1}{b-a}$. This means we need to multiply $(b-a)$ to the original sample average from simulation.  

The following is an illustration with estimation of $\theta = \int_{1.7}^{7.1} e^x dx$:
```{r answer}
n = 100000
a = 1.7
b = 7.1
U = runif(n, a, b)
thetaH = mean(exp(U))
thetaH #original sample average
(b-a)*thetaH #adjusted sample average
exp(b)-exp(a) #theoretical value
```